---
title: "What is R, and how do I use it?"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}

#load libraries
library(knitr)
library(learnr)
library(rmarkdown)
library(rsconnect)
library(tidyverse)
library(gapminder)
library(png)
#global chunk option
#knitr::opts_chunk$set(echo = FALSE)

#global objects
df <- gapminder |>
  filter(continent == "Africa") |>
  mutate(log_gdp = log10(gdpPercap)) |>
  arrange(desc(log_gdp))

df2 <- gapminder |>
  mutate(log_gdp = log10(gdpPercap))
```

## Hello, R!

Welcome to this brief introductory tutorial on using R. By the end of this lesson you will understand:

-   Why R is a useful tool
-   Some examples from the real world
-   How to install and load a package
-   How to do a common data manipulation
-   How to make a common plot

One important note. Do not worry if you don't understand how the functions work. Think of baking a cake. Do you understand how baking soda works? How protein content affects the sponge? You might, but it doesn't really matter if you just follow the instructions.

## Why R is a useful tool?

A programming environment, such as R, offers some cool stuff.

-   R is open-source and free. It has a huge support community that is constantly de-bugging and creating new functionalities.

-   If you find an analysis or cool example online, the code is almost always included. The R community is all about sharing.

-   R was developed for statistical programming.

-   If you can imagine an analytic task, I bet you can implement it in R. There may even be a specific package with pre-built functions to make it very easy.

-   Analyses in R are transparent, easily shareable, and reproducible. Can you remember every step you did to visualize a dataset in Excel so that someone else could add to it?

-   R can read and write data in virtually any data format.

-   R can be used for any data science task: scraping websites, developing websites, making static or interactive charts, making dashboards, automating repetitive tasks, statistical computations, text analysis, querying databases, and many others.

-   R has a lot of inter-operability with other platforms.

### Learning check
```{r checkbox, echo = FALSE}
question_checkbox("Select all of the reasons why R is useful:"
                  , answer("It is expensive", correct = FALSE, message = "It is the opposite of expensive."), answer("It can read and write data in virtually any format", correct = TRUE), answer("It promotes transparency in the analytical process", correct = TRUE), answer("It loves me", correct = FALSE, message = "R loves no one."), answer("It was developed specifically for statistics", correct = TRUE), answer("It provides a lot of options for a data scientist", correct = TRUE)
, allow_retry = TRUE)
```


## A simple exercise


Now that we've seen some examples from R, let's take a look at how R works through a simple exercise.


*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

What did you notice? Much like your old TI-83, R is built to handle numbers and do math without any additional steps involved.

*type some text in quotes*

```{r text, exercise = TRUE}

```

What did you notice? It treats it as text. With R you can paste together text, parse text, edit text, and analyze text.

And it can do so much more!


## Let's try it out!

### Installing and loading our first package

Let's load a package that includes some data. Run the code in the next chunk.

```{r load-gapminder, exercise = TRUE}

library(gapminder)
```

So, what just happened? We loaded the gapminder package which is a subset of the gapminder dataset.

**gapminder:** Gapminder is a country-level time-series dataset of many indicators that we are familiar with in the development field. The full dataset is available at https://www.gapminder.org/. 

Let's take a quick peak at the dataset.

```{r gapminder, exercise = TRUE, exercise.lines = 8}
#the function head shows the first 6 rows of the dataset

head(gapminder)

#the function tail shows the last 6 rows of the dataset

tail(gapminder)

glimpse(gapminder)
```


Now, it's your turn to load a suite of packages called the `tidyverse`. This suite contains the packages `dplyr` and `ggplot2` -- two packages that you'll use all the time in R for data analysis and visualization.

-   `dplyr` is a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges.
-   `ggplot2` is a system for declaratively creating graphics, based on The Grammar of Graphics. You provide the data, tell ggplot2 how to map variables to aesthetics, what graphical primitives to use, and it takes care of the details.

**Load the `tidyverse`**

```{r load-tidyverse, exercise = TRUE}



```

### Manipulate the gapminder

Now, let's use R to manipulate the gapminder dataset.

We are going to use three functions from `dplyr`.

* `filter()`: similar to using a filter on a data column in Excel.
* `mutate()`: adds a new column to the dataset. This is commonly used for adding a percentage, rescaling a variable, making a calculation using two or more columns, etc.
* `arrange()`: as the name applies it arranges the data by the column passed to the function. The default is ascending order. The exercise below shows how to use `arrange(desc())` to switch to descending order.

```{r gapminder_mani, exercise=TRUE, exercise.lines = 10, exercise.eval=TRUE}

df <- gapminder |>
  filter(continent == "Africa")|>   #filter the dataset
  mutate(log_gdp = log10(gdpPercap)) |> #add a new column
  arrange(desc(log_gdp)) #sort the data

#look at the data
head(df)

```

### Plot the data
The most common data visualization package is `ggplot2`. The `ggplot2` package is based on a specific grammar of graphics. Under this grammar of graphics, a visualization comprises six layers:

* data
* mapping (aesthetic)
* geometry
* facet
* statistic
* theme

All plots require the first three layers:

* data to visualize
* a mapping of variables from the data to the visualization grid
* different types of mappings are referred to as 'aesthetics'
* a geometry that gives a specific implementation of the specified aesthetic

ggplot(data = [DATASET],`
       mapping = aes(x = [X VARIABLE],`
                     y = [Y VARIABLE])) +`
  geom_SOMETHING()`

Below is some code to plot two of the variables in a scatterplot using `ggplot2`. In this example, 

* data = df
* mapping = aes(x = lifeExp, y = gdpPercap)
* geometry = geom_point() (produces a scatterplot)

We made the size of of each point equal to the population varialbe, and we made the color equal the name of each country. 

```{r gapminder_plot, exercise=TRUE,  out.width="75%", fig.align = "center"}


ggplot(df, aes(x = lifeExp, y = gdpPercap, size = pop)) +
  geom_point(aes(color = country)
             , alpha = .6) +
  theme(legend.position = "none")

```

Remember how we created a new variable called log_gdp? Recreate the above scatterplot using this new varialbe and compare the outputs.

```{r gapminder_v2, exercise = TRUE, exercise.lines = 10, out.width="75%", fig.align="center"}

```


```{r gapminder_v2-hint}

"copy and paste the code above and replace the y variable with log_gdp"

```
## Time to fly

Let's take advantage of the power of coding using our gapminder dataset. We made a single plot above, but what if we wanted to compare the logged gdps of countries over time for the whole world?

*First, add the log_gdp column to the df dataset and give it the name df2. *

```{r df_two, exercise = TRUE}


```

```{r df_two-hint}
"Look back at the previous exercise and use  mutate() to add the column"

```

```{r df_two-solution}
df2 <- gapminder |>
  mutate(log_gdp=log10(gdpPercap))

```


Now that we've created the object, we need to plot it.

```{r df_two_plot, exercise = TRUE, exercise.lines = 10}


```

```{r df_two_plot-hint}
"Use the previous ggplot code from above and change the y variable"

```

```{r df_two_plot-solution}

ggplot(df2, aes(x = lifeExp, y = log_gdp, color = continent, size = pop)) +
  geom_point(alpha = .6)

```

Wow, that's a lot of datapoints. Let's make one chart for each continent using ggplot's `facet_wrap` function.

* `facet_wrap()`: This function creates multiple plots using the variable that you pass to it.  

```{r df_two_facet, exercise = TRUE, exercise.lines = 10}


```

```{r df_two_facet-hint}

"ggplot uses the + operator to add lines to the code. Copy the plot used in the previous exercise and add a + before a new line for facet_wrap()."

"facet_wrap requires one argument, facets = 'object$column name'. You can specify how many rows or columns you want with nrow = [variable] and ncol = [variable]"

```

```{r df_two_facet-solution}

ggplot(df2, aes(x = lifeExp, y = log_gdp, color = continent, size = pop)) +
  geom_point(alpha = .6) +
  facet_wrap(facets = df2$continent
             , nrow = 2
             , ncol = 3)

```